\section{Income and consumption in the UK: data and measurement}\label{sec:measure}

In this section, we describe how we construct thirty years of micro-data measuring income and consumption for the same households. 

%As background, we then show how income and consumption give different impressions of the level and distribution of household resources in the UK, and how this changes when we include or do not include the imputed consumption or income from housing.

\subsection{the Living Costs and Food Survey}

Our main dataset is the Living Costs and Food Survey (known between 2001 and 2007 as the Expenditure and Food Survey, and the Family Expenditure Survey before that; we refer to it as the LCFS), the only UK survey which contains comprehensive measures of both income and expenditure. The LCFS is an annual, nationally-representative, cross-sectional survey with an annual sample of between 5,000 and 7,000 households. Its primary purpose is to provide data for the calculation of the commodity  weights for the UK's price indices. It aims to collect a comprehensive measure of household spending with a two-week diary, in which respondents are asked to record everything they  purchase, supplemented by a questionnaire in which respondents are asked about any spending on infrequently purchased items over the past number of months (for example, respondents are asked to record any spending on motor vehicles in the past 12 months, but any spending on household fuel in the past 3 months). It also collects data on household income measured over a relatively short period, as is standard in UK household surveys. We use data from 1978 to 2009.

\subsection{Income}
We use two different measures of income in this paper.  Our first is the usual measures of HBAI income before housing costs (BHC) (see Appendix \ref{data_annex} and DWP (XXXX)). The second is a broader measure of income intended to be directly comparable with our measure of consumption. To derive this, we start with the BHC income measure and subtract payments made to students from the Student Loan Company (these are loans, but, for reasons unclear to us, are treated like income in the HBAI income measure), and then add an estimate of the consumption flow from housing and motor vehicles net of the cash payments made on the same. The second of these adjustments takes account of the fact that ownership of a particular durable can be considered to yield an imputed flow of income just as it can be considered to yield an flow of consumption benefits, and it means that we make exactly the same adjustments to the income measure that we make when moving from expenditure to consumption./footnote{This partly addresses the concern in \cite{Bavier2008} about some of the arguments made by Meyer and Sullivan. Bavier argues that one should not compare consumption only to the measure of income used in the official analysis of poverty but to the best measure of income that can be derived. As the derivation of a consumption measure typically starts with expenditure data and makes adjustments in keeping with theoretical and empirical evidence about how best that data can be used to predict deprivation, then the odds are stacked against income predicting living standards better than consumption unless a similar process is carried out to the income data.} 

\subsection{Consumption}
The existing literature has tended to use either non-durable consumption, and total consumption (which differs from non-durable consumption by including the consumption flow from owned durables, the most important of which are vehicles and property). We use two measures of consumption. Both begin with a measure of household expenditure that simply records all spending by a household in a given period; we use, where available, responses to the questionnaire about spending on infrequently purchased items instead of information from the diary  [MIKE ASKS Cormac: DOES THIS NEED EXPANDING ? WHAT ABOUT BENEFITS IN KIND LIKE FREE SCHOOL MEALS AND FREE TV LICENSES? Are these in expenditure? consumption? or income?]. To derive a measure of full consumption, we begin with this measure of expenditure, subtract spending on vehicles and on housing (i.e. mortgage interest payments, capital payments and rent), viewing these outlays as investments in durables), and add an imputed consumption value for these two items (see Appendix \ref{data_annex} for details).\footnote{A measure of consumption should exclude items of spending that are in fact investments, but there is not yet consistency in the literature over how to treat (for example) spending on medical care or education; we do not exclude spending on such items from our measure, but, with a universal health care system that is free at the point use, these a lot less important for our UK data than for data from the US.} Our second measure is identical except that it does not include the imputed consumption flow from housing: this is therefore close to, but not the same as, non-durable consumption, but we create this measure to isolate how significant is the inclusion of the consumption flow from housing. 

There is an ongoing debate about the accuracy of the data on expenditure traditionally collected from household surveys (and used to derive measures of household consumption). A substantial problem is that of a declining coverage rate, where coverage is defined as the implied economy-wide expenditure captured by the CE survey as a ratio of that in the National Accounts: see, for example, \citet{Barrettetal2014}. Parker et al. (2009) and Aguiar and Bils (2015) both present evidence that the CE survey is, over time, increasingly under-measuring the expenditure of high-income households, which has clear implications for the use of consumption data to measure inequality. Accordingly, Meyer and Sullivan (2013a,b) and Aguiar and Bils (2015) both offer a method of adjusting reported expenditure in the CE survey to correct for the under-reporting. (However, in the other direction, Bee et al. (2014) presents evidence that the coverage rate of the CE survey - the ratio of spending captured by the CE survey to equivalent categories of spending in national accounts - is both high and not falling for several important expenditure categories.) There is also a considerable focus on the specifics of the Consumer Expenditure (CE) survey instrument, which collects information on expenditure using both a diary (to collect expenditure over a 2 week period) and an interview, where households are asked to recall expenditure in various categories over the past month (see Carroll et al. (2015) for a comprehensive tour of the key issues). For example, Attanasio et al. (2007) and (2012) compare the diary and interview components of the CE survey, and conclude that data from the Interview component understates the rise in consumption inequality, but Bee et al. (2014) argues that using the Diary data alone is also likely to produce misleading assessment of inequality trends; accordingly, Browning and Crossley (2009) and Battistan and Padula (2016) suggest different ways of combining the information from the Diary and the Interview components to reduce measurement error. 

[Some of this is a cut and paste from BEOD]. These issues have been less thoroughly studied in the UK. The data on household expenditure in the UK's LCFS/FES comes mainly from a diary, supplemented with information from an interview about spending on infrequently purchased items; unlike the CE survey, these diary and interview components provide complementary information and so there is little controversy over which to prefer or how best to combine them. Similar to the CE survey, the implied economy-wide expenditure captured by the LCFS is considerably lower than that reported in National Accounts (see \citealp{Deaton2005}, \citealp{Attanasio2006}, \citealp{Crossley2010} and \citet{Barrettetal2014}). We therefore might be concerned that expenditure data suffers from
its own under-reporting issues. The LCFS is the only nationally-representative
survey that collects comprehensive data on household budgets and,
as such, it is hard to validate household reports. However, \citet{LeicesterOldfield2009}
compare the expenditure patterns reported in the LCFS with those observed
in a commercial data set (the Kantar UK Worldpanel, previously known
as the TNS UK Worldpanel). The latter data covers spending on food
and groceries and is widely used in industry for market research.
\citet{LeicesterOldfield2009} find that, for each type of family,
levels of reported expenditure in the LCFS are \textit{higher} than those
in the Worldpanel.  

[We can also assess the quality of the spending data in the LCFS by comparing the estimates of total household income and consumption implied by the microdata with those reported in the UK National Accounts. This confirms other studies findings (Deaton, 2005; Attanasio et al., 2006) that there is an increasing (and alarming) gap between spending captured in the LCFS and spending reported in the National Accounts. However, we show that, as in the US (see Meyer and Sullivan, 2010) those items which make up a large fraction of spending of low-spending households have relatively good coverage rates, suggesting that consumption may be a less good indicator of the living standards of those with high resources than it is for those with low resources.]

\subsection{Adjusting for price changes and household composition}
We express all financial values in 2009 pounds, and use price indices based on the RPI to achieve this. \footnote{[PLEASE COULD CORMAC REVIEW THE ACCURACY AND NEED FOR THIS FOOTNOTE?] MS (2009) pay particular attention to how the choice of deflator materially affects conclusions about trends in living standards towards the bottom of the distribution. The UK has two main official measures of price inflation: the Retail Prices Index and the Consumers Prices Index: these differ in their formula and the coverage (for a summary of the differences, see Office for National Statistics (2011) [AND SHOULD WE REFER TO PETER'S PIECE?]). There are a number of reasons for our use of the RPI and its variants, rather than the CPI. These include the fact that it has been existence for the entirety of the period we consider (unlike the CPI); the fact that its coverage is broader (in particular it includes housing, which is omitted from the CPI), and the fact that the official poverty analyses produced by the UK government still (at time of writing) use the RPI rather than the CPI.} We do not use the actual RPI to deflate all of our measures of household resources, but instead make slight adjustments to reflect that our different measures of income and consumption are constructed in different ways, and therefore measure different things. In particular, we deflate measures of HBAI income using an official variant of the RPI which does not include the level of local taxation or housing depreciation in the basket of goods, and we deflate measures of consumption and income that include imputed housing income or consumption with our own variant of the RPI which does not include mortgage interest payments in the basket, and instead does include rent, but weighted in keeping with the budget share of imputed rent, rather than actual spending on rent. 

We construct measures of income and consumption at the level of the household, and then adjust for household composition using the modified OECD scale; this means we assume the same equivalence scale is applicable for our four different measures of household resources. Following usual UK practice, we re-base so that a two-adult household has a weight of 1, meaning that the scale becomes 0.67 for a single adult, 0.33 for each extra adult or child aged 14 or more, and 0.2 for every child aged under 14. [Mike to Cormac: I don't think this is true in this paper]. We conduct all analysis at the level of the individual (including children), having assigned to each individual their household's equivalised household income or consumption; this is numerically equivalent to having household-level data on equivalised household resources and weighting all analyses by the number of people in the household.

Appendix \ref{subsec:aggregate_trends} gives an overview of how income and consumption give different impressions of trends in the level and distribution of household resources in the UK, and how this changes when we include or do not include the imputed consumption or income from housing.